Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-02-22 paper/manuscript.Rmd
rezfddf

# Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running [@Fish1993], while drowning is also possible [@Miller1986].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].

@@

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond_2016]. We present two
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance [@Jeffery_2007], and discuss why these swimming events
appear to be relatively commonplace in the population.

# Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer_2015] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer_2015]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller_1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud_1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

This trade-off can be understood as density-dependent habitat selection governed
by the Ideal Free Distribution [@Morris_1987; @Bradbury_2015]. Ideal Free
Distribution theory predicts a fitness equilibrium: when the density in a given
habitat patch has exceeded the optimum for fitness within that patch, animals
should relocate and settle new habitat patches, so that fitness is equal across
all patches [@Bradbury_2015]. Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns remains unknown,
but swimming between islands may in part be explained by forage limitation and
conspecific competition.

# Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
(Miller 2002).<!-- find doi for this article --> During calving, female caribou
are known to swim to small islands in freshwater lakes and give birth, or with their
calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline
habitat in summer is considered an effective anti-predator calving strategy by
allowing for swimming to islands or peninsulas as a predation avoidance flight
response [@Bergerud_1990]. The use of water, and by association occasionally
swimming, is therefore part of the fine-scale interactions between caribou and
their predators and the use of islands appears to be an effective anti-predator
strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears
(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
adult caribou is relatively rare [@Bastille_Rousseau_2016], so the potential
effect of predators on caribou swimming behaviour is limited to non-consumptive
effects associated with encounters and subsequent flight response. Adult females
with calves at heel may be an exception because calves are highly susceptible to
coyotes [@Rayl_2014], but the energetic costs for calves to swim in the ocean is
likely far greater than adults. As one might expect, none <!--confirm--> of the
86 swimming events we observed occurred in the weeks after calves were born,
suggesting swimming with in the ocean with a calf-at-heel is likely more risky
than other predator avoidance behaviours.

@@

# Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
ultimate explanation to a series of proximate observations, fine-scale
density-dependent habitat selection is theoretically possible
[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
habitat was strongest at low population densities, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit -->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean (e.g. @Jeffery_2007; @Ricca_2012]. <!-- add Miller 1995, Miller
2002 -->  Taken together, we suggest that swimming is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean, with predator avoidance as a
potential proximate explanation. Although our inference is limited to
observations, our observations suggest that islands represent discrete foraging
patches for terrestrial animals that could influence fitness via increased
foraging opportunities and safety from predators. We also suggest that
individuals vary in their evaluation of costs and benefits of movement between
these habitat patches.

>>>>

# Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running <!--[@Fish1993]-->, while drowning is also possible [@Miller1985].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012].

@@

Caribou may swim between islands in the ocean for similar reasons that drive
caribou swimming in freshwater [@Leblond2016]. We present two
explanations for these swimming events, forage limitation [@Miller2002] and
predator avoidance [@Jeffery2007], and discuss why these swimming events
appear to be relatively commonplace in the population.

# Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer2016] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer2016]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

This trade-off can be understood as density-dependent habitat selection governed
by the Ideal Free Distribution [@Morris1987; @Bradbury2015]. Ideal Free
Distribution theory predicts a fitness equilibrium: when the density in a given
habitat patch has exceeded the optimum for fitness within that patch, animals
should relocate and settle new habitat patches, so that fitness is equal across
all patches [@Bradbury2015]. Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns remains unknown,
but swimming between islands may in part be explained by forage limitation and
conspecific competition.

# Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
[@Miller2002]. During calving, female caribou are known to swim to small islands
in freshwater lakes and give birth, or with their calves after birth, to avoid
predation [@Bergerud1985]. The use of shoreline habitat in summer is considered
an effective anti-predator calving strategy by allowing for swimming to islands
or peninsulas as a predation avoidance flight response [@Bergerud1990]. The use
of water, and by association occasionally swimming, is therefore part of the
fine-scale interactions between caribou and their predators and the use of
islands appears to be an effective anti-predator strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears
(*Ursus americanus*) are not present in the Fogo Island archipelago, thus
leaving coyotes (*Canis latrans*) as the top predator. Predation by coyotes on
adult caribou is relatively rare [@BastilleRousseau2016], so the potential
effect of predators on caribou swimming behaviour is limited to non-consumptive
effects associated with encounters and subsequent flight response. Adult females
with calves at heel may be an exception because calves are highly susceptible to
coyotes [@Rayl2014], but the energetic costs for calves to swim in the ocean is
likely far greater than adults. As one might expect, none <!--confirm--> of the
86 swimming events we observed occurred in the weeks after calves were born,
suggesting swimming with in the ocean with a calf-at-heel is likely more risky
than other predator avoidance behaviours.

@@

# Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989 ; Newfoundland and
Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation to a series of proximate observations, fine-scale density-dependent
habitat selection is theoretically possible [@Webber2018]. Predator avoidance
could also contribute to caribou swimming behaviour, but given the presumed low
density of coyotes on Fogo Island and that predation primarily only affects
young calves, we suspect predator is a possible, but less likely explanation.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit -->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Taken
together, we suggest that swimming is likely more common than previously thought
for caribou living on oceanic islands. We posit forage limitation and the
associated density-dependent habitat selection is an ultimate explanation for
why caribou swimming in the ocean, with predator avoidance as a potential
proximate explanation. Although our inference is limited to observations, our
observations suggest that islands represent discrete foraging patches for
terrestrial animals that could influence fitness via increased foraging
opportunities and safety from predators. We also suggest that individuals vary
in their evaluation of costs and benefits of movement between these habitat
patches.
